### 2013-12-09 report/paper.tex
added report files

%\documentclass[aps,prrd,preprint]{revtex4}
\documentclass[aps,prd,final,onecolumn,a4paper,10pt]{revtex4}

% Note:  comment out one of the two documentclass commands, depending
% on whether you need the final or preprint version. It is most convenient
% to work in preprint mode, and switch to final only at the end.

\usepackage{graphicx}                            % Graphics package
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage[margin=1in]{geometry}

\DeclareMathOperator*{\argmin}{arg\,min}

\begin{document}

\title{6.878 Final Report}
\author{Dalesh Dharamshi and Iris Xu}
\date{\today}

\begin{abstract}
\noindent
  In the era of personalized medicine it is anticipated that an individual patient's health record will contain multiple molecular expression profiles (genomic, transcriptomic, proteomic, metabolomic, etc) that have been captured across multiple time points in various health states. The combination of this data is expected to provide a much deeper understanding of an individual’s healthy and diseased states.

  Accurate identification of changes in molecular expression patterns across time and linking them to potential health conditions is at the core of the effectiveness of personalized medicine. Additionally, a patient is likely to have more than one health condition being expressed at the same time.

  Unlike a test for differential gene expression in one or more conditions, the clinical setting described above has the additional complexity of analyzing a time course of multiple biological components sampled across disparate time points and health conditions. The challenge is providing a concise description of an individual’s state that can be easily analyzed in this clinical setting. We investigate a few aspects of this complexity in order to facilitate generating a clinical recommendation.

  While previous work has focused on monitoring expression of specific genes, our goal is to provide an overview of an individual’s health state by identifying, separating, and annotating related biological component groups that have similar changes in expression levels, potentially time-shifted indicating a potential causal relationship.

  We find that while our method provides visual evidence of effective separation, enrichment analysis shows no significant enrichment between groups of signals.

\end{abstract}

\maketitle

\pagestyle{myheadings}
\markboth{!!!?!}{6.878 Final Report}
\thispagestyle{empty}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

\subsection{Personal Genomics}
With the advent of efficient sequencing, the possibility of personalized medicine based on an inidividual's specific genetic composition and molecular phenotype becomes

\subsection{Snyder Dataset}
Snyder

Chen, et al. examined how to generate an integrative personal omics profile to examine how biological components change during healthy and diseased states and how to use this information to estimate disease risk and learn about diseased states. While

\subsection{Other time series}

\section{Methods and Results}

\subsection{Pre-processing}

\subsection{Alignment}

The goal of our project was to find genes with similar expression patterns, indicating potential biological relevance.
In order to first find a measure of similarity, we calculated a distance metric between series using alignment.
We wanted similar expression signals to have a smaller distance. In particular, we wanted to capture the fact that changes in expression of one gene may affect the expression of another gene, but in the future. Additionally, we wanted to capture that the upregulation of a gene may cause the downregulation of another gene.
\\

Each processed gene expression series was converted to a character string.
We set a threshold change level, initally at $10$ percent.
If the expression level increased by more than the threshold from the previous timepoint, we convert to an \verb!R! for Rise.
If it decreased by more than the threshold, we convert to a \verb!D! for Drop.
And if it did not change by more than the threshold level, we convert to \verb!S! for Steady. The first character is always \verb!S!.\\

For example, the sequence $10$, $100$, $8$, $11$, $10$ is converted to \verb!S!, \verb!D!, \verb!R!, \verb!S!. Sequences were aligned and scored using Needleman-Wunsch.
In order to potentially align signals that have similar but negative expressions of each other, we attempted to give a smaller mismatch penalty between \verb!R! and \verb!D!.
We also give a very large gap penalty for large differences in time.
While sequence alignment is polynomial, we aligned on the order of $10^5$ sequences, so we modified our inital sequence alignment implementation to include gap penalty lookup as a speedup.
Alignment was run on a combination of AWS nodes and the Broad cluster.

\subsection{Clustering}

Given the distance matrix generated by alignment, we were able to cluster more similar sequences.

\subsection{Enrichment Analysis}
After identifying clusters, we wanted a way to efficiently describe each cluster, as well as to measure the effectiveness of our alignment and clustering algorithm.
To analyze the efficacy of alignment in clustering, we used Fischer's Exact test for Gene Ontology annotations.
Gerald Quon provided us with an \verb!R! script and Gene Ontology data for annotation testing of our clusters.
While, visually, the clusters showed alignment, the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions.\\

Additionally, we used \verb!DAVID! Bioinformatics Database web service to find functional groups within each cluster, as a means of describing each cluster, with $kappa=70$ and $linkage = 0.50$. We were able to generate functional groups for several clusters.

\subsection{Visualization}
In order to make expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3! that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and 
\section{Future Goals}

\section{Commetary on experience}
Iris: I learned a lot on how to pull data and to compare results with the ones in a paper. I also learned a lot about

\section{Commentary on peer-review process}
Iris: I thought that it was good to get feedback, but at the same time, our proposal was missing a lot of detail on purpose because we were not sure at the time exactly on what we were going to do, and a lot of the feedback was on lack of detail, which was not that helpful. It was nice to see other people's ideas.

\section{Division of labor}

Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the processing for the GSE675 dataset.

\section{Acknowledgements}
Thanks to Gerald Quon, Manolis Kellis, and Maxim Wolf for their help on our project.

\begin{thebibliography} {0}

\bibitem{Snyder} Chen R, Mias G I, Li-Pook-Than J, Jiang L, et al (2012). Personal Omics profiling reveals dynamic molecular and medical phenotypes. Cell 148, 1293-1307.

\bibitem{Gerald} Gerald Quon.

\bibitem{DAVID} Huang DW, Sherman BT, Lempicki RA (2009). Systematic and integrative analysis of large gene lists using DAVID Bioinformatics Resources. Nature Protoc 4(1), 44-57.

\bibitem{webservice} Huang DW, Sherman BT, Lempicki RA (2009). Bioinformatics enrichment tools: paths toward the comprehensive functional analysis of large gene lists. Nucleic Acids Res. 37(1), 1-13.

\end{thebibliography}

\end{document}

### 2013-12-09 ProjectCode_v1/submit_main.sh

@@ -1,8 +1,17 @@
 #!/bin/sh
-clusters="0 1"
+classes="const";
+#cell_types="GM12878 HELA K562 H1";
+cell_types="GM12878";
+#types="vanilla matchnodes matchdists matchboth";
+#nums="100 500 1000 2000";
+types="matchdists";             # Only 1 FG/BG match-type, to simplify to get Iris started.
+nums="1000";                    # Only one value for the number of interactions to be selected.
+sets="1 2 3 4 5"
+
+clusters = "0 1"
 
 for cluster in $clusters; do
   bsub -P compbiofolk -q compbio-week -J alignment_${cluster} \
-       -o out/output_alignment_${cluster} -n 8 \
-       -R "rusage[mem=16]" -R "span[hosts=1]" python Main.py ./Input/GSE675.csv GSE675_${cluster} ${cluster}
+       -o out/output_alignment_${cluster} -n 1 \
+       -R "rusage[mem=16]" python Main.py ./Input/mRNAExpression_FilteredBy5FPKM_QuantileNormalized.csv 5FPKMNormalizedAligned_${cluster} ${cluster}
 done;

### 2013-12-09 ProjectCode_v1/PreProcessAndUtils/ConvertCuffLinksOutputTocsv.py
New trials with GDS4238 dataset and full normalization of original dataset

@@ -1,0 +1,238 @@
+import sys
+import os
+
+TRANSCRIPT_FILE_NAME = "transcripts.gtf"
+ISOFORMS_FILE_NAME = "isoforms.fpkm_tracking"
+GENES_FILE_NAME = "genes.fpkm_tracking"
+INDEX_PREFIX = "GSM8185"
+FilterFPKMValue = 0.000
+# given a directory search for all transcripts.gtf files
+# read each transcripts.gtf file and create a matrix of genes vs fpkm values.
+
+
+
+#########################################################################
+##### getTranscriptFilesDictionary
+#########################################################################
+# searches for transcripts.gtf and creates a dictionary with the GSM** 
+#   number as the key and filename as value.
+# Input : directory to search
+# output : dictionary as described above.
+#########################################################################
+def getFilesDictionary(inputDirectory,fileNameType):
+    print " Processing ",inputDirectory,"...."
+    transcriptFilesDictionary = {}
+    for root, dirs, files in os.walk(inputDirectory):
+        for file in files:
+            if file.endswith(fileNameType):
+                fileName = os.path.join(root, file)
+                pos = fileName.find(INDEX_PREFIX)
+                indexKey = fileName[pos:pos+9]
+                transcriptFilesDictionary[indexKey] = fileName
+    return transcriptFilesDictionary
+
+#########################################################################
+##### fetchmRNAExpressionForFile
+#########################################################################
+# runs through each line looking for transcript entries and fills the rna expression
+# data with FPKM values
+#########################################################################
+def fetchmRNAExpressionForTranscriptFile(fileName,mRNAExpressionData):
+    stream = open(fileName)
+    for line in stream:
+        tokens = line.split()
+        if(tokens[2] == 'transcript'):
+            geneId = tokens[9][1:len(tokens[9])-2]
+            transcriptId=tokens[11][1:len(tokens[11])-2]
+            geneTranscriptId = geneId+":"+transcriptId
+            FPKMValue = float(tokens[13][1:len(tokens[13])-2])
+            if(mRNAExpressionData.has_key(geneTranscriptId)):
[…]

### 2013-12-09 ProjectCode_v1/PreProcessAndUtils/Log2RatioAndVectorNormalization.py

@@ -1,0 +1,40 @@
+import InputUtils
+import OutputUtils
+import sys
+import math
+
+def main():
+    if len(sys.argv) < 3:
+        print "you must call program as:  "
+        print "   pythonFPKMFiltering.py expressionRawFileName normColumn OutputFileName eg .\input\mRNAExpression_FilteredBy5FPKM_QuantileNormalized.csv  7 .\input\mRNAExpression_FilteredBy5FPKM_QuantileNormalizedLog2Filtered.csv "
+        return
+    rawExpressionFileName = sys.argv[1]
+    Signals,SignalSeqs,ExprNoToExprIdentifier,timeline = InputUtils.readExpressionTimeSeries(rawExpressionFileName)
+    normColumn = int(sys.argv[2])-1
+    mRNAExpressionDatacorrectedAndNormalized = {}
+    for key in ExprNoToExprIdentifier:
+        identifier  = ExprNoToExprIdentifier[key]
+        normValue = Signals[key][normColumn]
+        mRNAExpressionDatacorrectedAndNormalized[identifier] = Signals[key]
+        for i in xrange(len(Signals[key])):
+            newValue = math.log((Signals[key][i]+0.000000000001)/(normValue+0.000000000001),2)
+            mRNAExpressionDatacorrectedAndNormalized[identifier][i] = newValue
+    
+    # vector normalization.....
+    for j in xrange(len(mRNAExpressionDatacorrectedAndNormalized[ExprNoToExprIdentifier[0]])):
+        sumSq = 0.0
+        for key in ExprNoToExprIdentifier:
+            i  = ExprNoToExprIdentifier[key]
+            sumSq = sumSq + mRNAExpressionDatacorrectedAndNormalized[i][j]*mRNAExpressionDatacorrectedAndNormalized[i][j]
+        sqSumSq = math.sqrt(sumSq)
+        if(sqSumSq == 0):
+            sqSumSq = sys.float_info.epsilon
+        for key in ExprNoToExprIdentifier:
+            i  = ExprNoToExprIdentifier[key]
+            mRNAExpressionDatacorrectedAndNormalized[i][j] = mRNAExpressionDatacorrectedAndNormalized[i][j]/sqSumSq
+
+
+    OutputFileName = sys.argv[3]
+    OutputUtils.writeRNAExpressionasCSV(OutputFileName,mRNAExpressionDatacorrectedAndNormalized)
+    return
+main()

### 2013-12-09 report/paper.tex
more stuff for report

@@
 \subsection{Personal Genomics}
-With the advent of efficient sequencing, the possibility of personalized medicine based on an inidividual's specific genetic composition and molecular phenotype becomes
+With the advent of efficient sequencing techniques and improved computation resources, the possibility of personalized medicine based on an inidividual's specific genetic composition and molecular phenotype has increased dramatically over the past few years.
 
 \subsection{Snyder Dataset}
-Snyder
+The original dataset consisted of a genomic profile generated from blood samples taken from a 54-year-old male at 20 time points over the course of $401$ days.
+The supplemented the genomic data with blood glucose sampling, and increased the sampling rate during a human rhinovirus infection and a respiratory syncytial virus infection.
 
@@
 examined how to generate an integrative personal omics profile to examine how biological components change during healthy and diseased states and how to use this information to estimate disease risk and learn about diseased states.
-While
+Their analysis included whole genome sequencing, complete transcriptome analysis, proteomic and metabolomic analyses, and autoantibody profiles.
+Their multi-dimensional view of the individual showed how traits associated with disease could be monitored to identify physiological states.
+In particular, they processed the various omics sets using spectral analysis, clustering, and pathway enrichment analysis.
 
 \subsection{Other time series}
+In theory, our algorithms can work on any expression time series.
+
+We also tested our process on other pre-processed time series for gene expression including \verb!GSE675!, for HCMV infection of foreskin fibroblasts.
 
@@
 
+\begin{center}
+\begin{tabular}{c | ccc}
+  & R & D & S\\ \hline
+
+  R &  3 & -1 & -3\\
+  D & -1 & 3 & -3\\
+  S & -3 & -3 & 1\\
+\end{tabular}
+\quad
+\begin{tabular}{c | ccc}
+  & R & D & S\\ \hline
+
+  R &  3 & -3 & -1\\
+  D & -3 & 3 & -1\\
+  S & -1 & -1 & 3\\
+\end{tabular}
+\quad
+\begin{tabular}{c | ccc}
+  & R & D & S\\ \hline
+
+  R &  3 & -3 & -3\\
+  D & -3 & 3 & -3\\
+  S & -3 & -3 & 3\\
+\end{tabular}
+\end{center}
+
 \subsection{Clustering}
@@
 We were able to generate functional groups for several clusters.
+The \verb!DAVID!
+clustering provided functional groups within each cluster, whereas corrected Fischer's Exact test tries to summarize an entire cluster.
+Since we were able to find significant groups within each cluster, but not for an entire cluster, this may indicate that our clustering is too general or that the clusters are too large.
+
+\begin{figure}[h]
+\includegraphics[height=3in]{dendrogram.png}
+\caption{Dendrogram describing a cluster using our visualization tool.}
+\label{fig:dendrogram}
+\end{figure}
 
@@
 In order to make expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3!
-that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and
+that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and word clouds for a text summary of \verb!DAVID!
+gene groups by cluster.
 \section{Future Goals}
 
+We could integrate this with the SMART Genomics API to make our tool usable for all time series of gene expression.
 \section{Commetary on experience}
@@
 
+\bibitem{GSE675} Browne EP, Wing B, Coleman D, Shenk T.
+Altered cellular mRNA levels in human cytomegalovirus-infected fibroblasts: viral block to the accumulation of antiviral mRNAs.
+J Virol 2001 Dec;75(24):12319-30.
+PMID: 11711622
+
 \end{thebibliography}

### 2013-12-09 report/paper.tex
report changes.

@@
 % to work in preprint mode, and switch to final only at the end.
-
+\usepackage{float}
 \usepackage{graphicx}                            % Graphics package
@@
 \title{6.878 Final Report}
-\author{Dalesh Dharamshi and Iris Xu}
+\author{Iris Xu , Dalesh S Dharamshi }	
 \date{\today}
@@
 \noindent
-  In the era of personalized medicine it is anticipated that an individual patient's health record will contain multiple molecular expression profiles (genomic, transcriptomic, proteomic, metabolomic, etc) that have been captured across multiple time points in various health states.
-The combination of this data is expected to provide a much deeper understanding of an individual’s healthy and diseased states.
+  In the era of personalized medicine it is anticipated that an individual patient's health record will contain multiple molecular expression profiles that have been captured across multiple time points in various health states.
+Accurate identification \& representation of changes in molecular expression patterns across time and linking them to potential health conditions is at the core of the effectiveness of personalized medicine.
 
-  Accurate identification of changes in molecular expression patterns across time and linking them to potential health conditions is at the core of the effectiveness of personalized medicine.
-Additionally, a patient is likely to have more than one health condition being expressed at the same time.
+  We introduce sequence alignment score driven clustering to accurately identify changes in molecular expression over time , this approach can group time-shifted expression signals that may have a causal relationship.We further provide a concept of visualizing these clusters of change in molecular expression across time along with annotations of various health and biologically relevant terms enriched for each cluster.
 
-  Unlike a test for differential gene expression in one or more conditions, the clinical setting described above has the additional complexity of analyzing a time course of multiple biological components sampled across disparate time points and health conditions.
-The challenge is providing a concise description of an individual’s state that can be easily analyzed in this clinical setting.
-We investigate a few aspects of this complexity in order to facilitate generating a clinical recommendation.
+  Our results indicate that using alignment score as a distance to clustering produces clusters that includes signals that may have time shifted due to an upstream or downstream relation.These signals tend to be separated into different clusters while using the standard UPGMA based clustering.
 
-  While previous work has focused on monitoring expression of specific genes, our goal is to provide an overview of an individual’s health state by identifying, separating, and annotating related biological component groups that have similar changes in expression levels, potentially time-shifted indicating a potential causal relationship.
-
-  We find that while our method provides visual evidence of effective separation, enrichment analysis shows no significant enrichment between groups of signals.
+  Of the two time course datasets we tried this approach, only one of them shows a corresponding enrichment of biological terms associated identified clusters.
+The root cause for the failure in the other dataset is yet to be identified.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
 
-\subsection{Personal Genomics}
-With the advent of efficient sequencing techniques and improved computation resources, the possibility of personalized medicine based on an inidividual's specific genetic composition and molecular phenotype has increased dramatically over the past few years.
+\section{Background}
 
-\subsection{Snyder Dataset}
-The original dataset consisted of a genomic profile generated from blood samples taken from a 54-year-old male at 20 time points over the course of $401$ days.
-The supplemented the genomic data with blood glucose sampling, and increased the sampling rate during a human rhinovirus infection and a respiratory syncytial virus infection.
+	In the era of personalized medicine it is anticipated that an individual patient's health record will contain multiple molecular expression profiles that have been captured across multiple time points in various health states.
+Accurate identification \& representation of changes in molecular expression patterns across time and linking them to potential health conditions is at the core of the effectiveness of personalized medicine.Additionally, a patient is likely to have more than one health condition being expressed at the same time.
 
-Chen, et al.
-examined how to generate an integrative personal omics profile to examine how biological components change during healthy and diseased states and how to use this information to estimate disease risk and learn about diseased states.
-Their analysis included whole genome sequencing, complete transcriptome analysis, proteomic and metabolomic analyses, and autoantibody profiles.
-Their multi-dimensional view of the individual showed how traits associated with disease could be monitored to identify physiological states.
-In particular, they processed the various omics sets using spectral analysis, clustering, and pathway enrichment analysis.
+  Unlike a test for differential gene expression in one or more conditions, the clinical setting described above has the additional complexity of analyzing a time course of multiple biological components sampled across disparate time points and health conditions.
+The challenge is providing a concise description of an individual’s state that can be easily analyzed in this clinical setting.
+We investigate a few aspects of this complexity in order to facilitate generating a clinical recommendation.
 
-\subsection{Other time series}
-In theory, our algorithms can work on any expression time series.
+  While previous work has focused on monitoring and grouping molecular expression along the gene or conditions axes to identify changes , our goal is to introduce alignment to potentially improve the accuracy of this grouping by including potentially time-shifted signals that may have a relation.
+In addition we intend to arrive at an overview of the change in an individual’s health state by identifying, separating, and annotating related biological component groups that have similar changes in expression levels, potentially time-shifted indicating a potential causal relationship.
 
-We also tested our process on other pre-processed time series for gene expression including \verb!GSE675!, for HCMV infection of foreskin fibroblasts.
+  Note : While our overall project was done as part of two of our classes , in this report we highlight and detail the methods learnt and relevant to this class (Pre-processing , alignment , clustering).
+We only make a brief mention of visualization and annotation for completion.
 
-\section{Methods and Results}
+\section{Results and Discussion}
+\subsection{Input dataset and Preprocessing}
+\begin{description}
+ \item 1. GSE33029 : "Integrated analysis of omics profiles"
 
-\subsection{Pre-processing}
+ We have primarily used this ‘Snyderome’ dataset which contains various molecular expression captured across 20 time points over 14 months for a single individual.
+
+The RNASeq dataset posted on GEO is the output of topHat (~120GB) and we performed the following preprocessing.
+\begin{description}
+ \item a. Cufflinks analysis to derive gene expression abundance in FPKMs
+ \item b. Filtering and Quantile normalization of the cufflinks output.
+ \item c. We then created multiple views of this expression data to get a feel for how the
+		data looks.
+\end{description}
+
+In addition to the above we tested our methodology on the following two time course datasets.
+
+ \item 2. GSE19392 "Dynamic responses of primary human bronchial epithelial cells to influenza virus, viral RNA and interferon-beta"
+ \item 3. GSE675 Time course analysis of response to HCMV infection
+\end{description}
 
@@
 
-Given the distance matrix generated by alignment, we were able to cluster more similar sequences.
+Given the distance matrix generated by alignment, we were able to cluster similar sequences.
+Our clustering was based on the alignment scores generated above as distance.
+One question that needed to be resolved was how determine the optimal number of
+clusters.We have tried the following two approaches towards this
+a. Pairwise clustering until we reach a threshold of the alignment score between
+two clusters.
+b. Affinity propagation based clustering to automatically determine the number of
+clusters.The pairwise heirarchical clustering with an emperical value of threshold based
+out of experiments gives us fewer (about half) the clusters than the affinity propogation ,
+however the clusters generated by affinity propogation appear to be more cohesive
+
+Our first test is to check how the alignment and clustering compared to the
+preferred UPGMA clustering. We generated a heirarchical clustering output on the
+same preprocessed dataset with the help of an external tool ‘Cluster 3.0’
+developed by Michael Eisen at stanford.
+Shown below are three heatmaps the first two heatmaps are from clusters
+generated by our implementation of alignment and clustering (affinity prop
+followed by heirarchical) the third generated by cluster 3.0. In purple is an
+example of a cluster that based on the signal appears to belong together but is
+spread out in multiple clusters if alignment is not considered for clustering.
+
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.75]{ClustercomparisonResults_v2.png}
+\caption{Heatmap showing the comparison of aligned clustering and UPGMA.}
+\label{fig:ClusterComparison}
+\end{figure}
 

### 2013-12-09 report/paper.tex
revised paper

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{subcaption}
+
+\captionsetup{compatibility=false}
 
@@
 For example, the sequence $10$, $100$, $8$, $11$, $10$ is converted to \verb!S!, \verb!D!, \verb!R!, \verb!S!. Sequences were aligned and scored using Needleman-Wunsch.
-In order to potentially align signals that have similar but negative expressions of each other, we attempted to give a smaller mismatch penalty between \verb!R! and \verb!D!.
-We also give a very large gap penalty for large differences in time.
+In order to potentially align signals that have similar but negative expressions of each other, we attempted to give a smaller mismatch penalty between \verb!R!
+and \verb!D!.
+We also tried stricter versions of the scoring matrix to allow softened mismatch (between steady and anything else) or no mismatches (Table \ref({tab:scoring}).
+We also varied the gap penalty depending on how far apart the time points were. For very far apart signals, we used a larger penalty.
 While sequence alignment is polynomial, we aligned on the order of $10^5$ sequences, so we modified our inital sequence alignment implementation to include gap penalty lookup as a speedup.
@@
 
+\begin{table}
+\caption{Some sample scoring matrices alignment: least strict (allowing matches between drops and rises in signal for potention downregulation effects), more strict (giving mismatches between rises and drops), most strict (penalizing mismatches).
+The gap penalty depends on the time offset between two time points, increasing with time differences.}
+\label{tab:scoring}
 \begin{center}
@@
 \end{center}
+\end{table}
 
@@
 Gerald Quon provided us with an \verb!R! script and Gene Ontology data for annotation testing of our clusters.
-While, visually, the clusters showed alignment, the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions.\\
+While, visually, the clusters showed alignment, the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions with the Benjamini-Hochberg procedure.
+However, in order to get some quantitative measure of performance, we took sums of the p-value matrices generated from enrichment analysis for our technique for the GSE675 dataset and normalized over the same value for clusters generated by k-means (Fig.
+\ref({fig:enrichment}).
+It shows that, to some degree, our alignment and clustering technique performs better than without alignment.\\
 
-Additionally, we used \verb!DAVID!
+In order to provide a summarizing description of learned clusters we wanted to provide a word cloud, we initially pulled gene information from the \verb!OMIM!
+database.
+However, the text was too general even after multiple round of filtering and the generated word clouds were too non-descriptive.
+
+We then used \verb!DAVID!
 Bioinformatics Database web service to find functional groups within each cluster, as a means of describing each cluster, with $kappa=70$ and $linkage = 0.50$.
-We were able to generate functional groups for several clusters.
+We were able to generate functional groups for several clusters (Fig.
+\ref({fig:enrichment}).
 The \verb!DAVID!
@@
 \begin{figure}[h]
-\includegraphics[height=3in]{dendrogram.png}
-\caption{Dendrogram describing a cluster using our visualization tool.}
+
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \begin{subfigure}[b]{0.5\textwidth}
+    \includegraphics[width=.9\textwidth]{dendrogram.png}
     \label{fig:dendrogram}
+  \end{subfigure}%
+  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+    %(or a blank line to force the subfigure onto a new line)
+  \begin{subfigure}[b]{0.5\textwidth}
+    \includegraphics[width=.9\textwidth]{wordcloud.png}
+    \label{fig:wordcloud}
+  \end{subfigure}%
+  \label{fig:enrichment}
+
+  \begin{subfigure}[b]{0.8\textwidth}
+    \includegraphics[width=.7\textwidth]{pvalue_performace.png}
+    \label{fig:pvalue}
+  \end{subfigure}%
+  \label{fig:enrichment}
+  \caption{A) Dendrogram describing a cluster using our visualization tool.
+B) A word cloud generated from the clusters in the dendrogram.
+C) To give some performance metric, we took the ratio of the p-values for enrichment analysis for our technique over using plain k-means for the GSE675 dataset.}
 \end{figure}
@@
 that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and word clouds for a text summary of \verb!DAVID!
-gene groups by cluster.
+gene groups by cluster (Fig.
+\ref({fig:enrichment}).
 \section{Future Goals}
+We would re-do the pre-processing step or attempt to get the pre-processed dataset before reanalyzing our data.
 
 We could integrate this with the SMART Genomics API to make our tool usable for all time series of gene expression.
+
+\section{Comparison with original proposal} 
+We managed to complete everything that we proposed, including our alignment technique, several forms of clustering (Affinity Propogation, Hierarchical, K-means), several forms of enrichment analysis (DAVID's function annotations, pulling OMIM text, DAVID's gene grouping, Fischer's Exact test with multiple hypothesis corrections).
+We also provided a visualization tool as part of our combined class project.
+
 \section{Commetary on experience}
-Iris: I learned a lot on how to pull data and to compare results with the ones in a paper. I also learned a lot about
+Iris: I learned a lot on how to pull data and to compare results with the ones in a paper. I also learned a lot about pre-processing data and steps to take when results don't match those of the paper.
 
@@
 
-Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the processing for the GSE675 dataset.
+Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the analysis and processing for the GSE675 dataset.
 

### 2013-12-09 ProjectCode_v1/Alignment/seqalign.py

@@ -65,19 +65,6 @@ def seqalignDPFast(seq1,seq2,S,gap_pen,timelines,gapPenaltyLUT):
         F[0][j] = F[0][j-1] - gapPenaltyLUT[j-1][j]
     
     
-    
-    # YOUR CODE HERE
-    # Fill in the dynamic programming tables F and TB, starting at [1][1]
-    # Hints: The first row and first column of the table F[i][0] and F[0][j] are dummies
-    #        (see for illustration Durbin p.21, Figure 2.5, but be careful what you
-    #         think of as rows and what you think of as columns)
-    #        Hence, the bases corresponding to F[i][j] are actually seq1[i-1] and seq2[j-1].
-    #        Use the dictionary base_idx to convert from the character to an index to
-    #         look up entries of the substitution matrix.
-    #        To get started, you can complete and run the algorithm filling in only F,
-    #         and then figure out how to do TB.
-    
-    
     for i in range(1, len(seq1) + 1):
         for j in range(1, len(seq2) + 1):
             seq1Base = base_idx[seq1[i - 1]]

### 2013-12-09 ProjectCode_v1/Annotation/GeneClusters.py

@@ -39,11 +39,13 @@ def Geneclusters(clusterFolderName):
         print " NoOfEnsemble Symbols",len(EnsemblIdList)
         print "     Writing david file ..."
         #4. Generate reports from David.
-        
+        getDavidReports(GeneSymbolList,clusterName,clusterFolderName)
+        '''
         try:
             getDavidReports(EnsemblIdList,clusterName,clusterFolderName)
         except:
             print "Exception generating report for ",clusterName
+        '''
     f = open(clusterFolderName + "/geneCluster.json", 'w')
     jsonarray = json.dumps(geneClusters)
     f.write(jsonarray)
@@ -89,7 +91,7 @@ def getDavidReports(EnsemblIdList,clusterName,clusterFolderName):
     for i in xrange(len(EnsemblIdList)-1):
         inputIds = inputIds+EnsemblIdList[i]+","
     inputIds = inputIds+EnsemblIdList[len(EnsemblIdList)-1]
-    idType='ENSEMBL_GENE_ID'
+    idType='AFFYMETRIX_3PRIME_IVT_ID'
     listName = 'testList'
     listType = 0
     print "     Creating list of ",len(EnsemblIdList)," symbols"
@@ -103,7 +105,7 @@ def getDavidReports(EnsemblIdList,clusterName,clusterFolderName):
     initialSeed = 3
     finalSeed = 3
     linkage = 0.5
-    kappa = 70
+    kappa = 90
     geneClusterReportFile = open(clusterFolderName+"/"+clusterName+".GeneCluster","wb")
     geneClusterReport = client.service.getGeneClusterReport(overlap, initialSeed, finalSeed, linkage, kappa)
     #geneClusterReport = client.service.getGeneClusterReport(overlap, initialSeed, finalSeed, linkage, kappa)
@@ -180,7 +182,7 @@ def setup_logging():
         fmt = '%(asctime)s [%(levelname)s] %(funcName)s() @%(filename)s:%(lineno)d\n%(message)s\n'
     logging.basicConfig(level=logging.INFO, format=fmt)
 
-Geneclusters('../Output/0.05FPKMAgglomerativeClusters/OutputDirectoryHeirarchicalNoOfClusters_20')
+Geneclusters('../Output/GSE675_0/Heirarchical_NoOfClusters_9')
 
         
     

### 2013-12-09 ProjectCode_v1/submit_main.sh

@@ -1,17 +1,8 @@
 #!/bin/sh
-classes="const";
-#cell_types="GM12878 HELA K562 H1";
-cell_types="GM12878";
-#types="vanilla matchnodes matchdists matchboth";
-#nums="100 500 1000 2000";
-types="matchdists";             # Only 1 FG/BG match-type, to simplify to get Iris started.
-nums="1000";                    # Only one value for the number of interactions to be selected.
-sets="1 2 3 4 5"
-
-clusters = "0 1"
+clusters="0 1"
 
 for cluster in $clusters; do
   bsub -P compbiofolk -q compbio-week -J alignment_${cluster} \
-       -o out/output_alignment_${cluster} -n 1 \
-       -R "rusage[mem=16]" python Main.py ./Input/mRNAExpression_FilteredBy5FPKM_QuantileNormalized.csv 5FPKMNormalizedAligned_${cluster} ${cluster}
+       -o out/output_alignment_${cluster} -n 8 \
+       -R "rusage[mem=16]" -R "span[hosts=1]" python Main.py ./Input/GSE675.csv GSE675_${cluster} ${cluster}
 done;

### 2013-12-10 report/paper.tex
changes to the report.

@@
 
-  Of the two time course datasets we tried this approach, only one of them shows a corresponding enrichment of biological terms associated identified clusters.
-The root cause for the failure in the other dataset is yet to be identified.
+  Of the three time course datasets we tried this approach, only two of them appear to show a corresponding enrichment of biological terms associated with the identified clusters.
+The root cause for the failure for clusters to show significant enrichment in the third dataset is yet to be identified.
 
@@
 \section{Background}
+	\begin{figure}[H]
+\centering
+\includegraphics[scale=0.65]{Introduction.png}
+\caption{An individual patients molecular profile captured across time points.}
+\label{fig:PatientTimeCourses}
+\end{figure}
 
@@
   Note : While our overall project was done as part of two of our classes , in this report we highlight and detail the methods learnt and relevant to this class (Pre-processing , alignment , clustering).
-We only make a brief mention of visualization and annotation for completion.
+We make a brief mention of visualization and annotation for completion.
 
@@
  \item a. Cufflinks analysis to derive gene expression abundance in FPKMs
- \item b. Filtering and Quantile normalization of the cufflinks output.
+ \item b. Filtering,Quantile normalization,vector normalization of the cufflinks output.
  \item c. We then created multiple views of this expression data to get a feel for how the
@@
 
-In addition to the above we tested our methodology on the following two time course datasets.
+In addition to the above we tested our methodology on the following two time course gene expression datasets available on GEO. No additional preprocessing was done to these datasets.
 
@@
 clusters.The pairwise heirarchical clustering with an emperical value of threshold based
-out of experiments gives us fewer (about half) the clusters than the affinity propogation ,
-however the clusters generated by affinity propogation appear to be more cohesive
+out of experiments gives us fewer clusters than the affinity propogation . However the clusters 
+generated by affinity propogation appear to be more cohesive
 
-Our first test is to check how the alignment and clustering compared to the
-preferred UPGMA clustering. We generated a heirarchical clustering output on the
-same preprocessed dataset with the help of an external tool ‘Cluster 3.0’
-developed by Michael Eisen at stanford.
-Shown below are three heatmaps the first two heatmaps are from clusters
-generated by our implementation of alignment and clustering (affinity prop
-followed by heirarchical) the third generated by cluster 3.0. In purple is an
-example of a cluster that based on the signal appears to belong together but is
-spread out in multiple clusters if alignment is not considered for clustering.
+We compared our clustering and alignment to the standard UPGMA clustering.
+We generated a heirarchical clustering output on the same preprocessed dataset with the help of an external tool ‘Cluster 3.0’ developed by Michael Eisen at stanford.
+Shown below are three heatmaps the first two heatmaps are from clusters generated by our implementation of alignment and clustering (affinity prop followed by heirarchical) the third generated by cluster 3.0.
+In yellow is an example of a cluster that based on the signal appears to belong together but is spread out in multiple clusters if alignment is not considered for clustering.
 
@@
 After identifying clusters, we wanted a way to efficiently describe each cluster, as well as to measure the effectiveness of our alignment and clustering algorithm.
-To analyze the efficacy of alignment in clustering, we used Fischer's Exact test for Gene Ontology annotations.
+
+To analyze the efficacy of alignment in clustering, we used multiple approaches 1.
+Fischer's Exact test for Gene Ontology annotations.
+2.
+We ran the gene list generated by our clusters through DAVID functional annotation tool.
 Gerald Quon provided us with an \verb!R! script and Gene Ontology data for annotation testing of our clusters.
-While, visually, the clusters showed alignment, the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions with the Benjamini-Hochberg procedure.
+
+While, visually, the clusters showed alignment for GSE33029 dataset , the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions with the Benjamini-Hochberg procedure.
 However, in order to get some quantitative measure of performance, we took sums of the p-value matrices generated from enrichment analysis for our technique for the GSE675 dataset and normalized over the same value for clusters generated by k-means (Fig.
@@
 
+We have been unable to pinpoint the cause for the lack of enrichment annotations for our clusters on the GSE33029 dataset.We suspected that we are missing a noise correction step for the GSE33029 dataset that may be resulting in no enrichment showing up and reanalyzed with additional normalizations in no vain.
+We suspect the authors may have not fully specified the steps they took to preprocessing this dataset and we may be missing some correction step.
+
+To check if our approach works we decided to rerun our analysis on two additional and independent timecourse datasets from GEO (GSE19392 GSE675) .
+With clusters generated from these datasets we do significant enrichment numbers showing up for pathways and other annotation terms.
+The KEGG pathway results for a few of our clusters are as shown below.
+
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.65]{3Clusters_PathwayCombined.png}
+\caption{KEGG Pathway analysis on 3 clusters from GSE19392 dataset using david.}
+\label{fig:KEGGPathway.}
+\end{figure}
 \subsection{Visualization}
+
 In order to make expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3!
@@
 
+Dalesh: Working on the project was an enjoyable experience , a few highlights are as follows
+a. Opportunity to work in a team and have a mentor to guide us.
+b. Applying our learning on a couple of real datasets gave additional insights.
+c. The two interim feedback checkpoints
+
 \section{Commentary on peer-review process}
@@
 It was nice to see other people's ideas.
+Dalesh: The peer review process was fun .
+And i agree with Iris's comment on the feedback we received , i recollect giving ratings on the table without having completely understood what the authors were trying to communicate.
 
@@
 Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the analysis and processing for the GSE675 dataset.
+Dalesh: Most of the strategy and design was arrived at by brainstorming together , my specific implementation contributions were in data preprocessing on the amazon EC2 instances and clustering , I was responsible for processing and analyzing the GSE19392 dataset , we split the analysis of the GSE33029 dataset .
+Iris also did the implementations for the user interface.
 

### 2013-12-10 report/paper.tex
final submission

@@
 
-  Note : While our overall project was done as part of two of our classes , in this report we highlight and detail the methods learnt and relevant to this class (Pre-processing , alignment , clustering).
+  Note : While our overall project was done as part of two of our classes , in this report we highlight and detail the methods learnt and relevant to this class (Pre-processing , alignment , clustering, Enrichment analysis).
 We make a brief mention of visualization and annotation for completion.
@@
 
-The RNASeq dataset posted on GEO is the output of topHat (~120GB) and we performed the following preprocessing.
+The RNASeq dataset ( ~120 gb ) posted on GEO is the output of topHat and we had to perform the following preprocessing.
 \begin{description}
@@
 
-\begin{table}
+\begin{table}[H]
 \caption{Some sample scoring matrices alignment: least strict (allowing matches between drops and rises in signal for potention downregulation effects), more strict (giving mismatches between rises and drops), most strict (penalizing mismatches).
@@
 out of experiments gives us fewer clusters than the affinity propogation . However the clusters 
-generated by affinity propogation appear to be more cohesive
+generated by affinity propogation appear to be more cohesive.
 
@@
 
+We initially reused the pset versions of our alignment and clustering code but had to make more efficient versions to run through the larger dataset.
+
 \subsection{Enrichment Analysis}
@@
 While, visually, the clusters showed alignment for GSE33029 dataset , the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions with the Benjamini-Hochberg procedure.
-However, in order to get some quantitative measure of performance, we took sums of the p-value matrices generated from enrichment analysis for our technique for the GSE675 dataset and normalized over the same value for clusters generated by k-means (Fig.
-\ref({fig:enrichment}).
+However, in order to get some quantitative measure of performance, we took sums of the p-value matrices generated from enrichment analysis for our technique for the GSE675 dataset and normalized over the same value for clusters generated by k-means.
 It shows that, to some degree, our alignment and clustering technique performs better than without alignment.\\
+
+To further check if our approach works we decided to rerun our analysis on two additional and independent timecourse datasets from GEO (GSE19392 GSE675).
+With clusters generated from these datasets we do see significant enrichment numbers showing up for pathways and other annotation terms.
+The KEGG pathway results for a few of our clusters on the above datasets are as shown below.
+We havent yet related these results to the context of the original studies.
+Additional time is needed though to accurately conclude on the Enrichment analysis to pinpoint the cause for the lack of enrichment annotations for our clusters on the GSE33029 dataset.We suspected that we are missing a noise correction step for the GSE33029 dataset that may be resulting in no enrichment showing up and reanalyzed with additional normalizations in no vain.
+We also suspect the authors may have not fully specified the steps they took to preprocessing this dataset and we may be missing some correction step.
+
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.65]{3Clusters_PathwayCombined.png}
+\caption{KEGG Pathway analysis on 3 clusters from GSE19392 dataset using david.}
+\label{fig:KEGGPathway.}
+\end{figure}
 
@@
 Bioinformatics Database web service to find functional groups within each cluster, as a means of describing each cluster, with $kappa=70$ and $linkage = 0.50$.
-We were able to generate functional groups for several clusters (Fig.
-\ref({fig:enrichment}).
+We were able to generate functional groups for several clusters.
 The \verb!DAVID!
@@
 
-We have been unable to pinpoint the cause for the lack of enrichment annotations for our clusters on the GSE33029 dataset.We suspected that we are missing a noise correction step for the GSE33029 dataset that may be resulting in no enrichment showing up and reanalyzed with additional normalizations in no vain.
-We suspect the authors may have not fully specified the steps they took to preprocessing this dataset and we may be missing some correction step.
-
-To check if our approach works we decided to rerun our analysis on two additional and independent timecourse datasets from GEO (GSE19392 GSE675) .
-With clusters generated from these datasets we do significant enrichment numbers showing up for pathways and other annotation terms.
-The KEGG pathway results for a few of our clusters are as shown below.
-
-\begin{figure}[H]
-\centering
-\includegraphics[scale=0.65]{3Clusters_PathwayCombined.png}
-\caption{KEGG Pathway analysis on 3 clusters from GSE19392 dataset using david.}
-\label{fig:KEGGPathway.}
-\end{figure}
 \subsection{Visualization}
 
-In order to make expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3!
+In order to make the expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3!
 that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and word clouds for a text summary of \verb!DAVID!
-gene groups by cluster (Fig.
-\ref({fig:enrichment}).
+gene groups by cluster.
+
 \section{Future Goals}
-We would re-do the pre-processing step or attempt to get the pre-processed dataset before reanalyzing our data.
+We would redo the analysis on GSE33029 dataset , to identify the root cause of clusters not showing corresponding enrichment results.
 
@@
 \section{Comparison with original proposal} 
-We managed to complete everything that we proposed, including our alignment technique, several forms of clustering (Affinity Propogation, Hierarchical, K-means), several forms of enrichment analysis (DAVID's function annotations, pulling OMIM text, DAVID's gene grouping, Fischer's Exact test with multiple hypothesis corrections).
+Our original proposal had many items we were going to exlplore and choose from.
+We managed to complete everything that we proposed on our peer review response , including our alignment technique, several forms of clustering (Affinity Propogation, Hierarchical, K-means), several forms of enrichment analysis (DAVID's function annotations, pulling OMIM text, DAVID's gene grouping, Fischer's Exact test with multiple hypothesis corrections).
 We also provided a visualization tool as part of our combined class project.
@@
 b. Applying our learning on a couple of real datasets gave additional insights.
-c. The two interim feedback checkpoints
+c. Structuring of the project activity with milestones starting early in the class with two interim feedback checkpoints were helpful as well.
 
@@
 It was nice to see other people's ideas.
-Dalesh: The peer review process was fun .
-And i agree with Iris's comment on the feedback we received , i recollect giving ratings on the table without having completely understood what the authors were trying to communicate.
+
+Dalesh: The peer review discussion was fun but I agree with Iris's comment on the feedback we received , i recollect giving ratings on the table without having completely understood what the authors were trying to communicate and in many cases the initial proposals were in a premature stage.
 
@@
 Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the analysis and processing for the GSE675 dataset.
-Dalesh: Most of the strategy and design was arrived at by brainstorming together , my specific implementation contributions were in data preprocessing on the amazon EC2 instances and clustering , I was responsible for processing and analyzing the GSE19392 dataset , we split the analysis of the GSE33029 dataset .
+
+Dalesh: Most of the strategy and design was arrived at by brainstorming together , my specific implementation contributions were in data preprocessing on the amazon EC2 instances and implementations for clustering , I was responsible for processing and analyzing the GSE19392 dataset , we split the analysis of the GSE33029 dataset .
 Iris also did the implementations for the user interface.

### 2013-12-10 ProjectCode_v1/Alignment/AlignmentUtils.py
Added a project readme file and a few comments.

@@ -1,6 +1,8 @@
 import seqalign
 import timeit
 base_idx = { 'R' : 0, 'D' : 1, 'S' : 2}
+
+# Given a list of signals gets a distance matrix interms of the alignment score between them.
 def getAlignmentScoreMatrix(signalSeqs,S,gap_pen,timeline):
     scores = [[0] * len(signalSeqs) for i in range(len(signalSeqs))]
     startFull = timeit.default_timer()
@@ -16,7 +18,7 @@ def getAlignmentScoreMatrix(signalSeqs,S,gap_pen,timeline):
     return scores
     
 
-
+# Given a list of signals gets a distance matrix interms of the alignment score between them.
 def getAlignmentScoreMatrixWithLookup(signalSeqs,S,gap_pen,timeline):
     ScoreDictionary = {}
     lookupCount = 0
@@ -30,8 +32,8 @@ def getAlignmentScoreMatrixWithLookup(signalSeqs,S,gap_pen,timeline):
             doesScoreExist,lookupScore,lookupCount = lookUpScore(signalSeqs[i],signalSeqs[j],ScoreDictionary,lookupCount)
             if(doesScoreExist == True): scores[i][j] = lookupScore
             else:
-                scores[i][j] = seqalign.seqalignDP(signalSeqs[i],signalSeqs[j],S,gap_pen,timeline)
-                #scores[i][j] = seqalign.seqalignDPFast(signalSeqs[i],signalSeqs[j],S,gap_pen,timeline,gapPenLut)
+                scores[i][j] = seqalign.seqalignDPFast(signalSeqs[i],signalSeqs[j],S,gap_pen,timeline,gapPenLut)
+                #scores[i][j] = seqalign.seqalignDP(signalSeqs[i],signalSeqs[j],S,gap_pen,timeline)
                 addToLookupScore(signalSeqs[i],signalSeqs[j],scores[i][j],ScoreDictionary)
             scores[j][i] = scores[i][j]
     print " -- Successful lookups :",lookupCount
@@ -50,7 +52,8 @@ def getAlignmentScoreMatrixWithLookupPlusSignalKeys(signalSeqs,S,gap_pen,timelin
         for j in xrange(len(signalSeqs[i])):
             signalKey = signalKey*3+base_idx[signalSeqs[i][j]]
         seqKeys[i] = signalKey
-
+    
+    # lookup table for gap penalties.
[…]

### 2013-12-10 ProjectCode_v1/Annotation/OMIMAnnotation.py

@@ -1,6 +1,8 @@
 
 OMIMDict = {}
 
+# note that the omim.txt can be downloaded from omim and its too big to 
+# be placed in git.
 def GetOMIMDictionary():
     if(len(OMIMDict.keys()) > 2):
         return OMIMDict

### 2013-12-10 ProjectCode_v1/Clustering/AffinityPropogationClustering.py

@@ -2,6 +2,9 @@ from copy import deepcopy
 from sklearn.cluster import AffinityPropagation
 from numpy import matrix
 
+# there are two versions of affinity clustering in this code ,
+# the implemented version has some bugs so had to use one from sklearn.
+
 def AffinityPropCluster(S):
 
     SM = matrix(S)
@@ -24,7 +27,7 @@ def AffinityPropCluster(S):
     
     return clusters
 
-# this guy is sick....:-)..hence had to use the external
+# Some bug in this code , hence had to use an external affinity clustering 
 def AffinityPropClusterInternal(S):
     print "Indices :",cluster_centers_indices
     dampner = 0.5

### 2013-12-10 ProjectCode_v1/Clustering/Clustering.py

@@ -1,8 +1,6 @@
-
 import sys
 import csv
 from copy import deepcopy
-from sklearn.cluster import AffinityPropagation
 from numpy import matrix
 import  numpy as np
 import os
@@ -50,6 +48,7 @@ def initPairWiseAvgDistanceMatrix(ScoreMatrix):
         ADM[i,i] = VERY_LOW
     return ADM
 
+# ADM = average distance matrix.
 def getClustersWithMaxPairWiseAverageDistanceScoreFast(clusters,ScoreMatrix,ADM):
     rowLen = len(ADM[0:])
     maxindex = np.argmax(ADM)

### 2013-12-10 ProjectCode_v1/Main.py

@@ -107,24 +107,10 @@ def main():
     #OutputDirectoryName = "5FPKMNormalizedAlignedScoreMatrix"
     AlignAndCluster(Signals,SignalSeqs,ExprNoToExprIdentifier,S,gap_pen,timeline,OutputDirectoryName,ClusterType)
     #3. Annotation
-    #print "2. Perform Annotation ..."
-    #OutputDirectory = "./Output/"+OutputDirectoryName
-    # OutputDirs = [
-    #               '.\\Output\\0.05FPKMAgglomerativeClusters\\OutputDirectoryHeirarchicalNoOfClusters_80',
-    #               '.\\Output\\0.05FPKMAgglomerativeClusters\\OutputDirectoryHeirarchicalNoOfClusters_40',
-    #               '.\\Output\\0.05FPKMAgglomerativeClusters\\OutputDirectoryHeirarchicalNoOfClusters_30',
-    #               '.\\Output\\0.05FPKMAgglomerativeClusters\\OutputDirectoryHeirarchicalNoOfClusters_20',
-    #               '.\\Output\\0.05FPKMAgglomerativeClusters\\OutputDirectoryHeirarchicalNoOfClusters_10',
-    #               '.\\Output\\0.05FPKMAgglomerativeClusters\\OutputDirectoryHeirarchicalNoOfClusters_5',
-    #               '.\\Output\\0.05FPKMNormalizedAlignedAffinityPropogation',
-    #               '.\\Output\\5FPKMNormalizedAlignedAgglomerative\\OutputDirectoryHeirarchicalNoOfClusters_5',
-    #               '.\\Output\\0.05FPKMAgglomerativeClusters\\Heirarchical_NoOfClusters_2',
-    #               '.\\Output\\0.05FPKMAgglomerativeClusters\\Heirarchical_NoOfClusters_5',
-    #               '.\\Output\\0.05FPKMAgglomerativeClusters\\Heirarchical_NoOfClusters_10',
-    #               '.\\Output\\0.05FPKMAgglomerativeClusters\\Heirarchical_NoOfClusters_12'
-    #               ]
-    # for OutputDirectory in OutputDirs:
-    #     Annotation.Annotateclusters(OutputDirectory)
+    print "2. Perform Annotation ..."
+    OutputDirectory = "./Output/"+OutputDirectoryName
+    for OutputDirectory in OutputDirs:
+        Annotation.Annotateclusters(OutputDirectory)
     
     return
 

### 2013-12-10 report/paper.tex
updated paper

@@
   \begin{subfigure}[b]{0.5\textwidth}
+    \includegraphics[width=.9\textwidth]{graphs.png}
+    \label{fig:dendrogram}
+  \end{subfigure}%
+  ~
+  \begin{subfigure}[b]{0.5\textwidth}
     \includegraphics[width=.9\textwidth]{dendrogram.png}
@@
   \end{subfigure}%
-  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-    %(or a blank line to force the subfigure onto a new line)
+
   \begin{subfigure}[b]{0.5\textwidth}
-    \includegraphics[width=.9\textwidth]{wordcloud.png}
+    \includegraphics[width=.7\textwidth]{wordcloud.png}
     \label{fig:wordcloud}
   \end{subfigure}%
-  \label{fig:enrichment}
-
-  \begin{subfigure}[b]{0.8\textwidth}
-    \includegraphics[width=.7\textwidth]{pvalue_performace.png}
+  ~
+  \begin{subfigure}[b]{0.5\textwidth}
+    \includegraphics[width=.5\textwidth]{pvalue_performace.png}
     \label{fig:pvalue}
@@
   \label{fig:enrichment}
-  \caption{A) Dendrogram describing a cluster using our visualization tool.
-B) A word cloud generated from the clusters in the dendrogram.
-C) To give some performance metric, we took the ratio of the p-values for enrichment analysis for our technique over using plain k-means for the GSE675 dataset.}
+  \caption{A) The visualizer for gene expression.
+B) Dendrogram describing a cluster using our visualization tool.
+C) A word cloud generated from the clusters in the dendrogram.
+D) To give some performance metric, we took the ratio of the p-values for enrichment analysis for our technique over using plain k-means for the GSE675 dataset.}
 \end{figure}
@@
 
-\section{Comparison with original proposal} 
-Our original proposal had many items we were going to exlplore and choose from.
-We managed to complete everything that we proposed on our peer review response , including our alignment technique, several forms of clustering (Affinity Propogation, Hierarchical, K-means), several forms of enrichment analysis (DAVID's function annotations, pulling OMIM text, DAVID's gene grouping, Fischer's Exact test with multiple hypothesis corrections).
-We also provided a visualization tool as part of our combined class project.
-
-\section{Commetary on experience}
-Iris: I learned a lot on how to pull data and to compare results with the ones in a paper. I also learned a lot about pre-processing data and steps to take when results don't match those of the paper.
-
-Dalesh: Working on the project was an enjoyable experience , a few highlights are as follows
-a. Opportunity to work in a team and have a mentor to guide us.
-b. Applying our learning on a couple of real datasets gave additional insights.
-c. Structuring of the project activity with milestones starting early in the class with two interim feedback checkpoints were helpful as well.
-
-\section{Commentary on peer-review process}
-Iris: I thought that it was good to get feedback, but at the same time, our proposal was missing a lot of detail on purpose because we were not sure at the time exactly on what we were going to do, and a lot of the feedback was on lack of detail, which was not that helpful.
-It was nice to see other people's ideas.
-
-Dalesh: The peer review discussion was fun but I agree with Iris's comment on the feedback we received , i recollect giving ratings on the table without having completely understood what the authors were trying to communicate and in many cases the initial proposals were in a premature stage.
-
-\section{Division of labor}
-
-Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the analysis and processing for the GSE675 dataset.
-
-Dalesh: Most of the strategy and design was arrived at by brainstorming together , my specific implementation contributions were in data preprocessing on the amazon EC2 instances and implementations for clustering , I was responsible for processing and analyzing the GSE19392 dataset , we split the analysis of the GSE33029 dataset .
-Iris also did the implementations for the user interface.
-
 \section{Acknowledgements}
-Thanks to Gerald Quon, Manolis Kellis, and Maxim Wolf for their help on our project.
+Thanks to Peter Szolovits, Gil Alterovitz, and Gerald Quon for their help on our project.
 

### 2013-12-10 report/paper.tex
changed a bit
project report

@@
 
-\title{6.878 Final Report}
+\title{6.872 Final Report}
 \author{Iris Xu , Dalesh S Dharamshi }	
@@
 
-  We introduce sequence alignment score driven clustering to accurately identify changes in molecular expression over time , this approach can group time-shifted expression signals that may have a causal relationship.We further provide a concept of visualizing these clusters of change in molecular expression across time along with annotations of various health and biologically relevant terms enriched for each cluster.
+  We introduce sequence alignment score driven clustering to accurately identify changes in molecular expression over time , this approach can group time-shifted expression signals that may have a causal relationship.
+
+  We further provide a concept of visualizing these clusters of change in molecular expression across time along with annotations of various health and biologically relevant terms enriched for each cluster.
 
   Our results indicate that using alignment score as a distance to clustering produces clusters that includes signals that may have time shifted due to an upstream or downstream relation.These signals tend to be separated into different clusters while using the standard UPGMA based clustering.
-
   Of the three time course datasets we tried this approach, only two of them appear to show a corresponding enrichment of biological terms associated with the identified clusters.
 The root cause for the failure for clusters to show significant enrichment in the third dataset is yet to be identified.
+  Our visualzation provide an overview of changes represented by our gene clusters.
+However, fetching in real-time from DAVID web service, ENSEMBL, and OMIM is slow and makes real-time visualization of these annotations difficult.
 
@@
 \pagestyle{myheadings}
-\markboth{!!!?!}{6.878 Final Report}
+\markboth{!!!?!}{6.872 Final Report}
 \thispagestyle{empty}
@@
   While previous work has focused on monitoring and grouping molecular expression along the gene or conditions axes to identify changes , our goal is to introduce alignment to potentially improve the accuracy of this grouping by including potentially time-shifted signals that may have a relation.
-In addition we intend to arrive at an overview of the change in an individual’s health state by identifying, separating, and annotating related biological component groups that have similar changes in expression levels, potentially time-shifted indicating a potential causal relationship.
+  In addition we intend to arrive at a visual overview of the change in an individual’s health state by identifying, separating, and annotating related biological component groups that have similar changes in expression levels, potentially time-shifted indicating a potential causal relationship.
 
-  Note : While our overall project was done as part of two of our classes , in this report we highlight and detail the methods learnt and relevant to this class (Pre-processing , alignment , clustering, Enrichment analysis).
-We make a brief mention of visualization and annotation for completion.
+  Note : That we did our project as a combined work between two of our classes 6.872 and 6.878.
 
@@
 and \verb!D!.
-We also tried stricter versions of the scoring matrix to allow softened mismatch (between steady and anything else) or no mismatches (Table \ref({tab:scoring}).
+We also tried stricter versions of the scoring matrix to allow softened mismatch (between steady and anything else) or no mismatches (Table \ref{tab:scoring}).
 We also varied the gap penalty depending on how far apart the time points were. For very far apart signals, we used a larger penalty.
@@
 \end{figure}
-
-We initially reused the pset versions of our alignment and clustering code but had to make more efficient versions to run through the larger dataset.
 
@@
 
-In order to make the expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3!
+In order to make the expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool (Fig.4) using \verb!d3!
 that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and word clouds for a text summary of \verb!DAVID!
